\texttt{thornado}-transport: IMEX schemes for two-moment neutrino transport respecting Fermi-Dirac statistics

### 2019-01-31 Documents/M1/Astronum_2018/sections/TwoMoments.tex

\subsection{Transport Equations}
We consider neutrino transport through a static background, and include neutrino-matter interactions due to emission, absorption, and isoenergetic scattering.
After scaling to dimensionless units, the Boltzmann equation can be written as
\begin{equation}
  \pd{f}{t}+\vect{\ell}\cdot\nabla f
  =\f{1}{\tau}\,\cC(f),
  \label{eq:boltzmann}
\end{equation}
where the distribution function $f = f(\omega,\varepsilon,\vect{x},t)$ gives the number of neutrinos propagating in the direction $\omega\in\bbS^{2}$, with energy $\varepsilon\in\bbR^{+}$, at position $\vect{x}\in\bbR^{3}$ and time $t\in\bbR^{+}$.  
$\vect{\ell} = \vect{\ell}(\omega)\in\bbR^{3}$ is the unit vector parallel to the neutrino three-momentum vector: $\vect{p}=\varepsilon\,\vect{\ell}$.
On the right-hand side, $\tau$ is a collision time scale.  
In opaque regions, where neutrinos have frequent interactions with the background, $\tau\ll1$.  
In transparent regions, where neutrinos rarely interact and stream freely, $\tau\gg1$.
The collision term, $\cC(f)$, which models emission, absorption, and isoenergetic scattering is given by
\begin{equation}
  \cC(f)=\xi\,\big(\,f_{0}-f\,\big)
  +(1-\xi)\,\big(\,\f{1}{4\pi}\int_{\bbS^{2}}f\,d\omega-f\,\big),
  \label{eq:collisionTerm}
\end{equation}
where $\xi = \sigma_{\Ab} / (\sigma_{\Ab}  + \sigma_{\Scatt} )$ is the ratio of the absorption opacity $\sigma_{\Ab}$ to the total opacity.  
The scattering opacity is $\sigma_{\Scatt}$.  
The limit $\xi = 1$, when $\sigma_{\Scatt} = 0$, corresponds to pure absorption, while $\xi = 0$, when $\sigma_{\Ab} = 0$, corresponds to pure scattering.  
The equilibrium distribution function for neutrinos is given by the Fermi-Dirac distribution
\begin{equation}
  f_{0}(\vect{z})=\f{1}{e^{(\varepsilon-\mu(\vect{x}))/T(\vect{x})}+1}, 
  \label{eq:fermiDirac}
\end{equation}
where $\vect{z}:=\{\varepsilon,\vect{x}\}$, $T$ is the material temperature in energy units and $\mu$ is the neutrino chemical potential.
Both $T$ and $\mu$ depend on the spatial position $\vect{x}$.

@@

Approximate solutions to the Boltzmann equation, Eq.~\eqref{eq:boltzmann}, can be found by solving the two-moment model.
To this end, define the angular moments of the distribution function as follows
\begin{equation}
  \big\{\,\cJ,\vect{\cH},\vect{\cK}\,\big\}(\vect{z},t)
  =\f{1}{4\pi}\int_{\bbS^{2}}f(\omega,\vect{z},t)\,\{\,1,\vect{\ell},\vect{\ell}\otimes\vect{\ell}\,\}\,d\omega.
  \label{eq:angularMoments}
\end{equation}  
The zeroth moment, $\cJ$, is referred to as the particle density.  
The first moment, $\bcH$, is the particle flux, and the second moment, $\bcK$, is proportional to the stress tensor.  
By integrating Eq.~\eqref{eq:boltzmann} over the momentum space angular dimension we obtain equations for the zeroth and the first moments
\begin{equation}
  \pd{\vect{\cM}}{t}+\nabla\cdot\vect{\cF}=\f{1}{\tau}\,\vect{\cC}(\vect{\cM}),
  \label{eq:momentEquations}
\end{equation}
with $\vect{\cM}=(\cJ,\vect{\cH})^{T}$, $\vect{\cF}=(\vect{\cH},\vect{\cK})^{T}$, and
\begin{equation}
  \vect{\cC}(\vect{\cM})=\vect{\eta}-\vect{\cD}\,\vect{\cM},
  \label{eq:collisionTermMoments}
\end{equation}
where $\vect{\eta}=(\xi\,f_{0},\vect{0})^{T}$, $\vect{\cD}=\mbox{diag}(\xi,\vect{I})$, and
$\vect{I}$ is the identity matrix.
Hence, the process of solving the Boltzmann equation, Eq.~\eqref{eq:boltzmann}, for the neutrino distribution function $f(\omega,\vect{z},t)$, is replaced by solving the two-moment equations for the neutrino number density, $\cJ(\vect{z},t)$, and flux, $\bcH(\vect{z},t)$.

@@

As is shown in the previous section, the moment equation for $\bcH$ involves the higher moment $\bcK$ and the two-moment model is open.  
To close the two-moment model, we consider algebraic closures.  
For the two-moment model, algebraic closures give an approximation to $\bcK$ using the lower moments.  
Write
\begin{equation}
  \bcK = \vect{k} \cJ,
\end{equation}
where $\vect{k}$ is the Eddington tensor.  
By assuming that the distribution function is symmetric about a preferred direction $\widehat{\vect{h}}=\vect{\cH}/|\vect{\cH}|$, Levermore~\cite{levermore_1984} proposed a simple form for the Eddington tensor
\begin{equation}
  \vect{k}=\f{1}{2}\big[\,\big(1-\chi\big)\,\vect{I}+\big(3\,\chi-1\big)\,\widehat{\vect{h}}\otimes\widehat{\vect{h}}\,\big],
  \label{eq:eddingtonTensor}
\end{equation}
where $\chi=\chi(\cJ,|\vect{\cH}|)$ is the Eddington factor.  
Thus, the two-moment model is closed by specifying the scalar $\chi$ in terms of $\cJ$ and $|\vect{\cH}|$.

@@

Neutrinos are fermions and obey the Pauli exclusion principle.  
Because of this, the neutrino distribution function is bounded; i.e. $f \in [0,1]$.
As a result, the angular moments $\cJ$ and $\vect{\cH}$ and the Eddington factor $\chi$ satisfy the following bounds~\cite{levermore_1984,lareckiBanach_2011,kershaw_1976,shohatTamarkin_1943}: 
\begin{align}
\cJ \in[0,1], \quad &(1-\cJ)\cJ-|\vect{\cH}| \geq 0, \label{eq:MomentsBounds} \\
  \chi_{\mbox{\tiny min}}
  =\max\big(1-\f{2}{3\cJ},h^{2}\big)
  \leq & \chi \leq \min\big(1,\f{1}{3\cJ}-\f{\cJ}{1-\cJ}h^{2}\big)=\chi_{\mbox{\tiny max}},
  \label{eq:eddingtonFactorBounds}
\end{align}
where $h = |\bcH|/\cJ$ is the flux factor.

The constraints in Eq.~\eqref{eq:MomentsBounds} define realizable moments $\bcM$.  
For fermions, realizable moments can only be constructed from a distribution satisfying the bounds $f \in [0,1]$.  
Moreover, the set of realizable moments is convex: let $\cR$ contains all the possible $\bcM(f)$ so that for any  $f_1\in [0,1]$ and $f_2\in [0,1]$, $\bcM(f_1) \in \cR$ and $\bcM(f_2) \in \cR$, then $\lambda \bcM(f_1) + (1-\lambda)\bcM(f_2) \in \cR$ for any $\lambda \in [0,1]$~\cite{chu_etal_2018}.
As we will see later in Section~\ref{se:SpacialDiscretization}, this convexity makes it possible to design a realizability-preserving discretization for solving the two-moment model numerically.

The inequalities in Eq.~\eqref{eq:eddingtonFactorBounds} deserve further attention.  
They are equally important, relative to those in Eq.~\eqref{eq:MomentsBounds}, in maintaining consistency of the two-moment model with respect to Fermi-Dirac statistics.  
When designing a numerical scheme for the two-moment model that maintains realizable $\bcM$, it is also necessary for the Eddington factor to satisfy the bounds in Eq.~\eqref{eq:eddingtonFactorBounds}.

However, recently reported CCSN simulations using two-moment neutrino transport with algebraic closures have employed Eddington factors that can violate the bounds in Eq.~\eqref{eq:eddingtonFactorBounds}.  
As examples, we consider the Eddington factors discussed in~\cite{murchikova_etal_2017}, where the suitability of several algebraic closures for two-moment neutrino transport was evaluated.  
In Fig.~\ref{fig:EddingtonFactorsWithDifferentClosure}, we plot the Eddington factor versus the flux factor for two occupancies; $\cJ = 0.1$ (low occupancy) and $\cJ = 0.9$ (high occupancy).  
Of the algebraic closures plotted, few satisfy the bounds on the Eddington factor in Eq.~\eqref{eq:eddingtonFactorBounds}.  
Kershaw~\cite{kershaw_1976}, Wilson~\cite{wilson_1975,leblancWilson_1970}, Levermore~\cite{levermore_1984}, Minerbo~\cite{minerbo_1978}, and Janka~2~\cite{janka_1992} clsoures may work fine when the occupancy is low.  
When the occupancy is high, the Eddington factor due to these closures exceeds the upper bound for Fermi-Dirac statistics.  
The Eddington factor of Janka~1~\cite{janka_1991} may violate both the upper and lower bound on the Eddington factor.  
Only the closure due to Cernohorsky~\&~Bludman~\cite{cernohorskyBludman_1994} satisfies both the upper and lower bounds.  
This is not surprising as this is the only closure based on Fermi-Dirac statistics.  
Although the Levermore and Minerbo closures do not satisfy the bounds in Eq.~\eqref{eq:eddingtonFactorBounds}, they are widely used in simulations of neutrino transport in CCSNe and compact binary mergers; e.g., O'Connor \& Couch~\cite{oConnorCouch_2018}, Pan et al.~\cite{pan_etal_2018}, Glas et al.~\cite{glas_etal_2018}, Just et al.~\cite{just_etal_2018}, and Foucart et al.~\cite{foucart_etal_2015} use the Minerbo closure, while Vartanyan et al.~\cite{vartanyan_etal_2018}, Cabezon et al.~\cite{cabezon_etal_2018}, Kuroda et al.~\cite{kuroda_etal_2016}, and Fujibayashi et al.~\cite{fujibayashi_etal_2017} use the Levermore closure.  
When employing these closures in conditions of high occupancy, a numerical scheme may evolve the moments outside the realizable domain of Fermi-Dirac statistics given in Eq.~\eqref{eq:MomentsBounds}.
If this were to happen, the update step may give $f>1$, which would change the sign of the blocking factors in the collision terms.
If no corrective actions were taken, it would be difficult to predict the impact of the subsequent induced errors on the simulation outcome (later, we will encounter examples in the 
test cases we have considered).
If treatments are developed to map the unrealizable moments into the realizable domain, they must be developed to conserve lepton number, energy, and momentum.

>>>>

\subsection{Transport Equations}
We consider neutrino transport through a static background, and include neutrino-matter interactions due to emission, absorption, and isoenergetic scattering.
After scaling to dimensionless units, the Boltzmann equation can be written as
\begin{equation}
  \pd{f}{t}+\vect{\ell}\cdot\nabla f
  =\f{1}{\tau}\,\cC(f),
  \label{eq:boltzmann}
\end{equation}
where the distribution function $f = f(\omega,\varepsilon,\vect{x},t)$ gives the number of neutrinos propagating in the direction $\omega\in\bbS^{2}$, with energy $\varepsilon\in\bbR^{+}$, at position $\vect{x}\in\bbR^{3}$ and time $t\in\bbR^{+}$.  
$\vect{\ell}(\omega)\in\bbR^{3}$ is the unit vector parallel to the neutrino three-momentum: $\vect{p}=\varepsilon\,\vect{\ell}$.
On the right-hand side, $\tau$ is a collision time scale.  
In opaque regions, where neutrinos have frequent interactions with the background, $\tau\ll1$.  
In transparent regions, where neutrinos rarely interact and stream freely, $\tau\gg1$.
The collision term, $\cC(f)$, which models emission, absorption, and isoenergetic scattering is given by
\begin{equation}
  \cC(f)=\xi\,\big(\,f_{0}-f\,\big)
  +(1-\xi)\,\big(\,\f{1}{4\pi}\int_{\bbS^{2}}f\,d\omega-f\,\big),
  \label{eq:collisionTerm}
\end{equation}
where $\xi = \sigma_{\Ab} / (\sigma_{\Ab}  + \sigma_{\Scatt} )$ is the ratio of the absorption opacity $\sigma_{\Ab}$ to the total opacity.  
The scattering opacity is $\sigma_{\Scatt}$.  
The limit $\xi = 1$, when $\sigma_{\Scatt} = 0$, corresponds to pure absorption, while $\xi = 0$, when $\sigma_{\Ab} = 0$, corresponds to pure scattering.  
The equilibrium distribution function for neutrinos is given by the Fermi-Dirac distribution:
\begin{equation}
  f_{0}(\vect{z})=\f{1}{e^{(\varepsilon-\mu(\vect{x}))/T(\vect{x})}+1}, 
  \label{eq:fermiDirac}
\end{equation}
where $\vect{z}:=\{\varepsilon,\vect{x}\}$, $T$ is the material temperature in energy units and $\mu$ is the neutrino chemical potential.
Both $T$ and $\mu$ depend on the spatial position $\vect{x}$.

@@

Approximate solutions to the Boltzmann equation, Eq.~\eqref{eq:boltzmann}, can be found by solving the two-moment model.
To this end, define the angular moments of the distribution function as follows:
\begin{equation}
  \big\{\,\cJ,\vect{\cH},\vect{\cK}\,\big\}(\vect{z},t)
  =\f{1}{4\pi}\int_{\bbS^{2}}f(\omega,\vect{z},t)\,\{\,1,\vect{\ell},\vect{\ell}\otimes\vect{\ell}\,\}\,d\omega.
  \label{eq:angularMoments}
\end{equation}  
The zeroth moment, $\cJ$, is referred to as the particle density.  
The first moment, $\bcH$, is the particle flux, and the second moment, $\bcK$, is proportional to the stress tensor.  
By integrating Eq.~\eqref{eq:boltzmann} over the momentum-space angular dimension we obtain equations for the zeroth and the first moments:
\begin{equation}
  \pd{\vect{\cM}}{t}+\nabla\cdot\vect{\cF}=\f{1}{\tau}\,\vect{\cC}(\vect{\cM}),
  \label{eq:momentEquations}
\end{equation}
with $\vect{\cM}=(\cJ,\vect{\cH})^{T}$, $\vect{\cF}=(\vect{\cH},\vect{\cK})^{T}$, and
\begin{equation}
  \vect{\cC}(\vect{\cM})=\vect{\eta}-\vect{\cD}\,\vect{\cM},
  \label{eq:collisionTermMoments}
\end{equation}
where $\vect{\eta}=(\xi\,f_{0},\vect{0})^{T}$, $\vect{\cD}=\mbox{diag}(\xi,\vect{I})$, and
$\vect{I}$ is the identity matrix.
Hence, the process of solving the Boltzmann equation, Eq.~\eqref{eq:boltzmann}, for the neutrino distribution function $f(\omega,\vect{z},t)$, is replaced by solving the two-moment equations for the neutrino number density, $\cJ(\vect{z},t)$, and flux, $\bcH(\vect{z},t)$.

@@

The moment equation for $\bcH$ involves the higher moment $\bcK$ and the two-moment model is open.  
To close the two-moment model, we consider algebraic closures.  
For the two-moment model, algebraic closures give an approximation to $\bcK$ using the lower moments.  
Write
\begin{equation}
  \bcK = \vect{k} \cJ,
\end{equation}
where $\vect{k}$ is the Eddington tensor.  
By assuming that the distribution function is symmetric about a preferred direction $\widehat{\vect{h}}=\vect{\cH}/|\vect{\cH}|$, Levermore~\cite{levermore_1984} proposed a simple form for the Eddington tensor:
\begin{equation}
  \vect{k}=\f{1}{2}\big[\,\big(1-\chi\big)\,\vect{I}+\big(3\,\chi-1\big)\,\widehat{\vect{h}}\otimes\widehat{\vect{h}}\,\big],
  \label{eq:eddingtonTensor}
\end{equation}
where $\chi=\chi(\cJ,|\vect{\cH}|)$ is the Eddington factor.  
Thus, the two-moment model is closed by specifying the scalar $\chi$ in terms of $\cJ$ and $|\vect{\cH}|$.

@@

Neutrinos are fermions and obey the Pauli exclusion principle.  
Because of this, the neutrino distribution function is bounded; i.e.~$f \in [0,1]$.
As a result, the angular moments $\cJ$ and $\vect{\cH}$ and the Eddington factor $\chi$ satisfy the following bounds~\cite{levermore_1984,lareckiBanach_2011,kershaw_1976,shohatTamarkin_1943}: 
\begin{align}
\cJ \in[0,1], \quad &(1-\cJ)\cJ-|\vect{\cH}| \geq 0, \label{eq:MomentsBounds} \\
  \chi_{\mbox{\tiny min}}
  =\max\big(1-\f{2}{3\cJ},h^{2}\big)
  \leq & \chi \leq \min\big(1,\f{1}{3\cJ}-\f{\cJ}{1-\cJ}h^{2}\big)=\chi_{\mbox{\tiny max}},
  \label{eq:eddingtonFactorBounds}
\end{align}
where $h = |\bcH|/\cJ$ is the flux factor.

The constraints in Eq.~\eqref{eq:MomentsBounds} define realizable moments $\bcM$.  
For fermions, realizable moments can only be constructed from a distribution satisfying the bounds $f \in [0,1]$.  
Moreover, the set of realizable moments is convex: let $\cR$ be the realizability set and $\bcM_{1}, \bcM_{2} \in \cR$, then $\lambda \bcM_{1} + (1-\lambda)\bcM_{2} \in \cR$ for any $\lambda \in [0,1]$~\cite{chu_etal_2018}.
As we will see later in Section~\ref{se:SpatialDiscretization}, this convexity makes it possible to design a realizability-preserving discretization for solving the two-moment model numerically.

The inequalities in Eq.~\eqref{eq:eddingtonFactorBounds} deserve further attention.  
They are important as those in Eq.~\eqref{eq:MomentsBounds}, in maintaining consistency of the two-moment model with respect to Fermi-Dirac statistics.  
When designing a numerical scheme for the two-moment model that maintains realizable $\bcM$, it is also necessary for the Eddington factor to satisfy the bounds in Eq.~\eqref{eq:eddingtonFactorBounds}.

However, recently reported CCSN simulations using two-moment neutrino transport with algebraic closures have employed Eddington factors that can violate the bounds in Eq.~\eqref{eq:eddingtonFactorBounds}.  
As examples, we consider the Eddington factors discussed in~\cite{murchikova_etal_2017}, where the suitability of several algebraic closures for two-moment neutrino transport was evaluated.  
In Fig.~\ref{fig:EddingtonFactorsWithDifferentClosure}, we plot the Eddington factor versus the flux factor for two occupancies; $\cJ = 0.1$ (low occupancy) and $\cJ = 0.9$ (high occupancy).  
Of the algebraic closures plotted, few satisfy the bounds on the Eddington factor in Eq.~\eqref{eq:eddingtonFactorBounds}.  
Kershaw~\cite{kershaw_1976}, Wilson~\cite{wilson_1975,leblancWilson_1970}, Levermore~\cite{levermore_1984}, Minerbo~\cite{minerbo_1978}, and Janka~2~\cite{janka_1992} closures may work fine when the occupancy is low.  
When the occupancy is high, the Eddington factor due to these closures exceeds the upper bound for Fermi-Dirac statistics.  
The Eddington factor of Janka~1~\cite{janka_1991} may violate both the upper and lower bound on the Eddington factor.  
Only the closure due to Cernohorsky~\&~Bludman~\cite{cernohorskyBludman_1994} satisfies both the upper and lower bounds.  
This is not surprising as this is the only closure based on Fermi-Dirac statistics.  
Although the Levermore and Minerbo closures do not satisfy the bounds in Eq.~\eqref{eq:eddingtonFactorBounds}, they are widely used in simulations of neutrino transport in CCSNe and compact binary mergers; e.g., O'Connor \& Couch~\cite{oConnorCouch_2018}, Pan et al.~\cite{pan_etal_2018}, Glas et al.~\cite{glas_etal_2018}, Just et al.~\cite{just_etal_2018}, and Foucart et al.~\cite{foucart_etal_2015} use the Minerbo closure, while Vartanyan et al.~\cite{vartanyan_etal_2018}, Cabezon et al.~\cite{cabezon_etal_2018}, Kuroda et al.~\cite{kuroda_etal_2016}, and Fujibayashi et al.~\cite{fujibayashi_etal_2017} use the Levermore closure.  
When employing these closures in conditions of high occupancy, a numerical scheme may evolve the moments outside the realizable domain of Fermi-Dirac statistics given in Eq.~\eqref{eq:MomentsBounds}.
If this were to happen, the update step may give $\cJ>1$.
Considering the fact that the completed collision term contains blocking factors, i.e.~$(1-\cJ)\times$ something positive, $\cJ > 1$ would change blocking factors' sign and it would be difficult to predict the impact of the subsequent induced errors on the simulation outcome.
\rc{Besides, even with the simplified collision term, Eq~\ref{eq:collisionTermMoments}, $\cJ>1$ could result... the reason of crash}
If treatments are developed to map the unrealizable moments into the realizable domain, they should be developed to conserve lepton number, energy, and momentum.

### 2019-01-31 Documents/M1/Astronum_2018/sections/intro.tex

One essential part of the explosion mechanism is neutrino transport.
Neutrino energy deposition is believed to drive CCSN explosions except in cases where rapid rotation is present and magnetohydrodynamic effects may dominate 
(for a review, see~\cite{mezzacappa_2005,janka_2012,burrows_2013,muller_2016}).
Ideally, neutrino transport would be modeled by the Boltzmann transport equation, which is an integro-partial-differential equation evolving a phase-space distribution function $f$ (e.g., see~\cite{mezzacappaBruenn_1993a,mezzacappaBruenn_1993b,mezzacappaBruenn_1993c,mezzacappa_etal_2001,liebendoerfer_etal_2001,liebendoerfer_etal_2004,livne_etal_2004,liebendoerfer_etal_2005,ott_etal_2008,sumiyoshiYamada_2012,nagakura_etal_2014,sumiyoshi_etal_2014,nagakura_etal_2018}.
Simulating the neutrino transport implies finding a solution of the Boltzmann equation for a specific domain and period with acceptable accuracy.

However, solving the Boltzmann transport equation is at present too expensive.
To balance physical fidelity and computational expediency, an approximate method called the two-moment method has been adopted 
(e.g., see~\cite{kuroda_etal_2016,roberts_etal_2016,just_etal_2018,vartanyan_etal_2018}).
Using the two-moment method, the evolved variables are the zeroth and first angular moments of the distribution function $f$ -- the spectral particle density $\cJ$ and flux $\bcH$, respectively.
However, the transport term in the Boltzmann transport equation includes the dot product of the velocity vector and the gradient of the phase-space distribution function.
Integrating this term with an unit direction vector as the weight function couples the moment equations: it introduces the second angular moment $\bcK$ to the first moment $\bcH$'s equation.
Knowledge of $\bcK$ is needed to close the two-moment equation system.
Therefore, an algebraic closure that gives a predicted $\bcK$ given $\cJ$ and $\bcH$ is needed.
The better the closure predicts $\bcK$, the more accurate the two-moment method. 
The two-moment method has been widely applied with different algebraic closures, such as the Minerbo~\cite{minerbo_1978} closure(e.g.~{O'Connor} and {Couch}~\cite{oConnorCouch_2018}, Pan and et al.~\cite{pan_etal_2018}, Glas et al.~\cite{glas_etal_2018}, and Just et al.~\cite{just_etal_2018} ) and the Levermore~\cite{levermore_1984} closure(e.g.~Vartanyan et al.~\cite{vartanyan_etal_2018}, Cabezon et al.~\cite{cabezon_etal_2018}, and Kuroda et al.~\cite{kuroda_etal_2016}).

Applying the two-moment method does simplify the problem, but doesn't guarantee an affordable solution.
To be precise, how to discretize the continuous equation system given by the two-moment method and solve the discretized system efficiently remains a question.
In fact, the time scales of neutrino interactions with the background ($\sim\mathcal{O}(10^{-13})$~second) is short compared to the duration of the CCSN explosion ($\sim\mathcal{O}(1)$~second).  
This means that $\sim\mathcal{O}(10^{13})$ time steps would be needed for solving the system fully explicitly. 
On the other hand, solving the moment equations fully implicitly requires inverting globally many band-structured matrices whose sizes depend on the phase-space discretization.
Such a global inversion is both expensive and unfriendly to parallelization.
To circumvent these challenges, implicit-explicit (IMEX) methods are taken into consideration.
By treating the transport terms in the two-moment equations explicitly and the collision terms implicitly, IMEX methods are subject only to a time step governed by the explicit transport terms, and the matrices to be inverted are block diagonal.
Therefore, IMEX methods require far fewer time steps compared with a fully explicit method, and the computation for each step is easily parallelizable.  
%For a non-relativistic system where the propagating speed of gravitional wave and fluid is much slower than the light speed, fully implicit method is a better choice for neutrino transport in CCSN simulation.~\cite{cernohorskyVanWeert_1992}(read ?)
For the relativistic setting that we have, where the fluid and the neutrinos have relativistic propagation speeds, IMEX methods are efficient.

To model neutrino transport using a two-moment method, two things need to be chosen carefully: an algebraic closure based on Fermi-Dirac statistics for closing the two-moment equations and a convex-invariant, diffusion-accurate IMEX scheme to ensure a physical result.
A convex-invariant scheme is a scheme that has the property: if all the possible n-th step has $u^{n}\in W$ and $W$ is convex, then all the possible (n+1)-th step has $u^{n+1}\in W$.
Since the neutrino distribution function is bounded ($f\in[0,1]$) by the Pauli exclusion principle, its moments as weighted integrals of a bounded function over the domain $\omega\in\bbS^{2}$ are also bounded.
We call the moments satisfying the constraints due to Fermi-Dirac statistics \textit{realizable moments} and the realizable $\cJ$ and $\bcK$ define a convex set~\cite{chu_etal_2018}.
The algebraic closure should give a realizable $\bcK$, and the well-posedness of the closure requires realizable $\cJ$ and $\bcH$.
It explains why an algebraic closure based on Fermi-Dirac statistics is needed.
Realizability of $\cJ$ and $\bcH$ after each time step requires a convex-invariant IMEX scheme.
Since the realizable moments form a convex set, it is possible to construct a realizability-preserving method using a convex-invariant IMEX scheme for two-moment neutrino transport.
What's more, the physics of neutrino transport in CCSNe requires the IMEX scheme to be diffusion-accurate.

The study of moment realizability and realizability-preserving methods with diffusion-accurate IMEX schemes motivates this work.
Gottlieb et al.~\cite{gottlieb_etal_2001} showed that standard strong-stability-preserving IMEX scheme cannot have an order higher than first without a restricted time step.
To have second-order (or higher-order) accuracy, some correction steps are needed.
Unfortunately, the correction steps can deteriorate the accuracy of the IMEX scheme in the diffusion limit or restrict the time step.
To keep things simple, we focus on IMEX schemes without correction steps and require them to be high-order (second or higher order) in the streaming limit and diffusion-accurate.
We call these IMEX schemes \textit{PD-ARS}.

\texttt{thornado} is our toolkit for high-order neutrino-radiation hydrodynamics based on high-order Runge-Kutta Discontinuous Galerkin (RKDG) methods.
It is being developed at the University of Tennessee, Knoxville and Oak Ridge National Laboratory.
It currently solves the Euler equations for fluid dynamics and the two-moment approximation of the radiative transfer equation, both in the non-relativistic limit~\cite{endeve_etal_2018}.

This paper is organized as follows: Section~\ref{se:Two-MomentModel} discusses the mathematical model, algebraic closures, and the constraints on the moments and algebraic closures imposed by Fermi-Dirac statistics;
Section~\ref{se:SpacialDiscretization} gives a first-order finite-volume spatial discretization and shows how the spatial discretization preserves constraints in an IMEX step;
Section~\ref{se:TimeIntegration} discusses how to use convex combination to construct two PD-ARS schemes, one with second-order accuracy in the streaming limit and the other with third-order accuracy in the same limit;
Section~\ref{se:NumericalTests} presents the results of the numerical tests, which demonstrate the properties of the PD-ARS schemes; Section~\ref{se:Conclusion} summarizes the achievements of this paper and discusses future work.

%\rc{Reference check}

>>>>

One essential part of the explosion mechanism is neutrino transport.
Neutrino energy deposition is believed to be the major driver of CCSN explosions, -- except in cases where rapid rotation is present and magnetohydrodynamic effects may dominate (for reviews, see~\cite{mezzacappa_2005,janka_2012,burrows_2013,muller_2016}).
Ideally, neutrino transport would be modeled by the Boltzmann transport equation, which is an integro-partial-differential equation evolving a phase-space distribution function $f$.~(e.g., see~\cite{mezzacappaBruenn_1993a,mezzacappaBruenn_1993b,mezzacappaBruenn_1993c,mezzacappa_etal_2001,liebendoerfer_etal_2001,liebendoerfer_etal_2004,livne_etal_2004,liebendoerfer_etal_2005,ott_etal_2008,sumiyoshiYamada_2012,nagakura_etal_2014,sumiyoshi_etal_2014,nagakura_etal_2018} for studies of CCSN with Boltzmann transport in various approximate settings.)
Simulating the neutrino transport implies finding a solution of the Boltzmann equation for a specific domain and period, with acceptable accuracy.

However, solving the Boltzmann transport equation with sufficient phase-space resolution and full weak interaction physics is at present too expensive.
To balance physical fidelity and computational expediency, an approximate method called the two-moment method has been adopted (e.g., see~\cite{kuroda_etal_2016,roberts_etal_2016,just_etal_2018,vartanyan_etal_2018}).
Using the two-moment method, the evolved variables are the zeroth and first angular moments of the distribution function $f$ -- the spectral particle density $\cJ$ and flux $\bcH$, respectively.
However, the transport term in the Boltzmann transport equation includes the dot product of the velocity vector and the gradient of the phase-space distribution function.
Integrating this term with a unit direction vector as the weight function couples the moment equations: it introduces the second angular moment $\bcK$ to the first moment $\bcH$'s equation.
Knowledge of $\bcK$ is needed to close the two-moment system.
Therefore, a closure that gives $\bcK$ consistent with $\cJ$ and $\bcH$ is needed.
The better the closure predicts $\bcK$, the more accurate the two-moment method. 
The two-moment method has been widely applied in the CCSN modeling community with different algebraic closures, such as the Minerbo~\cite{minerbo_1978} closure (e.g.~{O'Connor} and {Couch}~\cite{oConnorCouch_2018}, Pan and et al.~\cite{pan_etal_2018}, Glas et al.~\cite{glas_etal_2018}, and Just et al.~\cite{just_etal_2018}) and the Levermore~\cite{levermore_1984} closure (e.g.~Vartanyan et al.~\cite{vartanyan_etal_2018}, Cabezon et al.~\cite{cabezon_etal_2018}, and Kuroda et al.~\cite{kuroda_etal_2016}).

Applying the two-moment method does simplify the problem, but doesn't guarantee an affordable solution.
To be precise, how to discretize the continuous equation of system given by the two-moment method and solve the discretized system efficiently remains a question.
In fact, the time scales of neutrino interactions with the background (can be $\sim\mathcal{O}(10^{-13})$~second) is short compared to the duration of the CCSN explosion ($\sim\mathcal{O}(1)$~second).  
This means that $\sim\mathcal{O}(10^{13})$ time steps would be needed for solving the system fully explicitly. 
On the other hand, solving the moment equations fully implicitly requires inverting global band-structured matrices whose sizes depend on the phase-space discretization.
Such a global inversion is both expensive and unfriendly to parallelization.
To circumvent these challenges, implicit-explicit (IMEX) methods are taken into consideration.
By treating the transport terms in the two-moment equations explicitly and the collision terms implicitly, IMEX methods are subject only to a time step governed by the explicit transport terms, and the matrices to be inverted are block diagonal.
Therefore, IMEX methods require far fewer time steps compared with a fully explicit method, and the computation for each step is easily parallelizable.
For the relativistic setting that we have, where the fluid and the neutrinos have comparable propagation speeds, and IMEX methods can be efficient.

To model neutrino transport using a two-moment method, two (or at least two) things need to be chosen carefully: an algebraic closure based on Fermi-Dirac statistics for closing the two-moment equations and a convex-invariant, diffusion-accurate IMEX scheme to ensure a physical result.
A convex-invariant scheme has the following property: if the solution $u^{n}\in W$ and $W$ is convex, then $u^{n+1}\in W$.
Since the neutrino distribution function is bounded ($f\in[0,1]$) by the Pauli exclusion principle, its moments as weighted integrals of a bounded function over the domain $\omega\in\bbS^{2}$ are also bounded.
We call the moments satisfying the constraints due to Fermi-Dirac statistics \textit{realizable moments} and the realizable $\cJ$ and $\bcH$ define a convex set~\cite{chu_etal_2018}.
The algebraic closure should give a realizable $\bcK$, and the well-posedness of the closure requires realizable $\cJ$ and $\bcH$.
It explains why an algebraic closure based on Fermi-Dirac statistics is needed.
Realizability of $\cJ$ and $\bcH$ after each time step requires a convex-invariant IMEX scheme.
Since the realizable moments form a convex set, it is possible to construct a realizability-preserving method using a convex-invariant IMEX scheme for two-moment neutrino transport.
In addition, the physics of neutrino transport in CCSNe requires the IMEX scheme to be diffusion-accurate.

The study of moment realizability and realizability-preserving methods with diffusion-accurate IMEX schemes motivates this work.
Gottlieb et al.~\cite{gottlieb_etal_2001} showed that standard strong-stability-preserving IMEX schemes cannot have an order higher than first without a restricted time step.
To have second-order (or higher-order) accuracy, some correction steps are needed.
Unfortunately, the correction steps can deteriorate the accuracy of the IMEX scheme in the diffusion limit or restrict the time step.
To keep things simple, we focus on IMEX schemes without correction steps and require them to be high-order (second or higher order) in the streaming limit and diffusion-accurate.
We call these IMEX schemes \textit{PD-ARS}.

\texttt{thornado} is our toolkit for high-order neutrino-radiation hydrodynamics based on high-order Runge-Kutta Discontinuous Galerkin (RKDG) methods.
It is being developed at the University of Tennessee, Knoxville and Oak Ridge National Laboratory.
It currently includes solvers for the Euler equations for fluid dynamics and the two-moment approximation of the radiative transfer equation~\cite{endeve_etal_2018}.
In this paper, we focus on the transport methods in \texttt{thornado} with emphasis on IMEX.

This paper is organized as follows: Section~\ref{se:Two-MomentModel} discusses the mathematical model, algebraic closures, and the constraints on the moments and algebraic closures imposed by Fermi-Dirac statistics;
Section~\ref{se:SpatialDiscretization} gives a first-order finite-volume spatial discretization and shows how the spatial discretization preserves constraints in an IMEX step;
Section~\ref{se:TimeIntegration} discusses how to use convex combination to construct two PD-ARS schemes, one with second-order accuracy in the streaming limit and the other with third-order accuracy in the same limit;
Section~\ref{se:NumericalTests} presents the results of the numerical tests, which demonstrate the properties of the PD-ARS schemes; Section~\ref{se:Summary} summarizes the achievements of this paper and discusses future work.
